Parsing Postfix log files

### 2009-02-12 doc/logparser-implementation.tex
Fixed acronyms so they work properly, get expanded when appropriate, and add the     correct entries to the glossary and acronym appendices. Correct capitalisation of most section and subsection headings. Corrections made to background chapter; more remain to be made.

\subsection{Rules table}

@@

\subsection{Connections table}

@@

    \item [server\_ip] The \gls{IP} address (IPv4 or IPv6) of the server:
        the local address when receiving mail, the remote address when
        sending mail.

    \item [server\_hostname] The hostname of the server, it will be
        \texttt{unknown} if the \gls{IP} address could not be resolved to a
        hostname via DNS\@.

    \item [client\_ip] The client \gls{IP} address (IPv4 or IPv6): the
        remote address when receiving mail, the local address when sending
        mail.

    \item [client\_hostname] The hostname of the client, it will be
        \texttt{unknown} if the \gls{IP} address could not be resolved to a
        hostname via DNS\@.

    \item [helo] The hostname used in the HELO command.  The HELO hostname
        occasionally changes during a connection, presumably because spam
        or virus senders think it is a good idea.  By default Postfix only
        logs the HELO hostname when it rejects an \gls{SMTP} command, but
        it is quite simple to rectify this, as described in
        \sectionref{logging helo}.

@@

\subsection{Results table}

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \gls{SMTP} command ---
        a mechanism that is quite useful for testing new restrictions.
        This field will be 1 if the log line parsed was a warning rather
        than a real rejection, or 0 for a real rejection.  This field
        should be ignored if the result is not a rejection, i.e.\ the
        action field of the associated rule is not \texttt{REJECTION}.

    \item [smtp\_code] The \gls{SMTP} code associated with the log line.
        In general an \gls{SMTP} code is only present in a log line for a
        rejection or final delivery; results initially missing an
        \gls{SMTP} code will duplicate the \gls{SMTP} code of other results
        in that connection.  Some final delivery log lines do not contain
        an \gls{SMTP} code: in those cases the code is specified by the
        rule, based on the success or failure represented by the log line.

@@

    \item [data] A field available for anything not covered by other
        fields, e.g.\ the rejection message from a \gls{DNSBL}\@.

@@

\subsection{Example rule}

@@

\subsection{Creating new rules}

@@

The log files produced by Postfix differ from installation to installation,
because administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different \gls{DNSBL}
services, policy servers, or custom rejection messages.  To facilitate
parsing new log lines, the parser's design separates parsing rules from
parsing actions: adding new actions can be difficult, but adding new rules
to parse new rejection messages is trivial, and occurs much more
frequently.  A program is provided to ease the process of creating new
rules from unparsed log lines, based on the algorithm developed by Risto
Vaarandi for his \gls{SLCT}~\cite{slct-paper}.  The differences between the
two algorithms will be outlined as part of the general explanation below.

The core of the \gls{SLCT} algorithm is quite simple: log lines are
generally created by substituting variable words into a fixed pattern, and
analysis of the frequency with which each word occurs can be used to
determine whether the word is variable or part of the fixed pattern.  This
classification can be used to group similar log lines and generate a regex
to match each group of log lines.

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained when writing rules and performs a large number of
        substitutions on the input log lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \gls{IP} addresses, the
        standard start of rejection messages, etc.) with regex keywords
        that the parser will expand when it loads the rule (see the
        regex entry in \sectionref{rule attributes}).  The purpose of
        this step is to utilise existing knowledge to create more accurate
        regexes; it replaces a large number of variable words with fixed
        words, improving the subsequent classification of words as fixed or
        variable.  The new log lines are written to a temporary file, which
        all subsequent stages use instead of the original input file.

@@

\subsection{Actions in detail}

@@

            \item [pid] The \gls{pid} of the program that logged the line.

@@

    \item [CONNECT] Handle a remote client connecting: create a new state
        table entry (indexed by \daemon{smtpd} \gls{pid}) and save both the
        client hostname and \gls{IP} address.

@@

    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an \gls{SMTP}
        command from the remote client: log the rejection with a mail if
        there is a queueid in the log line, or with the connection if not.

@@

\subsection{Postfix actions}

@@

\subsection{Adding new actions}

@@

    \item Mail enters the system via \gls{SMTP} or local submission.

@@

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop}, by \gls{SMTP}, by re-injection due to
forwarding, or internally generated by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately and the sender address is
logged (action: PICKUP\@; flowchart:~2).  Re-injection due to forwarding
sadly lacks explicit log lines of its own; it is explained in
\sectionref{Re-injected mails}.

@@

\gls{SMTP} is more complicated than local submission:

@@

    \item This is followed by rejection of sender address, recipient
        addresses, client \gls{IP} address or hostname, HELO hostname,
        etc.\ (action: REJECTION\@; flowchart:~4); acceptance of one or
        more mails (action: CLONE\@; flowchart:~5); or some interleaving of
        both.

    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
        Postfix has rejected any \gls{SMTP} commands the data will be saved
        to the database; if not there will not be any data to save (any
        mails accepted will already have been cloned so their data is in
        another data structure).

@@

\section{Complications encountered}

@@

\subsection{Queueid vs pid}

The mail lacks a queueid until it has been accepted, so log lines must
first be correlated by the \daemon{smtpd} \gls{pid}, then transition to
being correlated by the queueid.  This is relatively minor, but does
require:

@@

    \item Most importantly: every section of code must know whether it
        needs to lookup the data structures by \gls{pid} or queueid.

@@

\subsection{Connection reuse}

@@

Multiple independent mails may be delivered across one connection: this
requires the algorithm to clone the current data as soon as a mail is
accepted, so that subsequent mails will not trample over each other's data.
This must be done every time a mail is accepted, as it is impossible to
tell in advance which connections will accept multiple mails.  Once the
mail has been accepted its log lines will not be correlated by \gls{pid}
any more, its queueid will be used instead (except when timeouts occur
during the data phase \sectionref{timeouts-during-data-phase}).  If the
original connection has any useful data (e.g.\ rejections) it will be saved
to the database when the client disconnects.  One unsolved difficulty is
distinguishing between different groups of rejections, e.g.\ when dealing
with the following sequence:

@@

    \item The client issues the RSET command to reset the \gls{SMTP}
        session.

@@

\subsection{Re-injected mails}

@@

    \item Local delivery re-injects the mail and logs a relayed delivery
        rather than delivering directly to a mailbox or program as it
        usually would.\footnote{Relayed delivery is performed by the
        \gls{SMTP} client; local delivery means local to the server, i.e.\
        an address the server is final destination for.}  In this case the
        mail may already have been created (described above) and the
        unknown origin flag will be cleared; if not a new data structure
        will be created.  In both cases the re-injected mail is marked as a
        child of the original mail.  The log line in question is:

@@

\subsection{Identifying bounce notifications}

@@

    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
        messages associated with the mail, indicating it was generated
        internally by Postfix, not accepted via \gls{SMTP} or submitted
        locally by \daemon{postdrop}.

@@

\subsection{Aborted delivery attempts}

@@

Some mail clients behave unexpectedly during the \gls{SMTP} dialogue: the
client aborts the first delivery attempt after the first recipient is
accepted, then makes a second delivery attempt for the same recipient which
it continues with until delivery is complete.  Microsoft Outlook is one
client that behaves in this fashion; other clients may act in a similar
way.  An example dialogue exhibiting this behaviour is presented below
(lines starting with a three digit number are sent by the server, the other
lines are sent by the client):

@@

If all the checks above are successful then the mail is assumed to be an
aborted delivery attempt and is discarded.  There will be no further
entries logged for such mails, so without identifying and discarding them
they accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and \gls{IP} address, but the database
schema requires many more fields be populated (see \sectionref{connections
table} and \sectionref{results table}).  This heuristic is quite
restrictive, and it appears there is little scope for false positives; if
there are any false positives there will be warnings when the next log line
for that mail is parsed.  False negatives are less likely to be detected:
there may be queueid clashes (and warnings) if mails remain in the state
tables after they should have been removed, otherwise the only way to
detect false negatives is to examine the state tables after each parsing
run.

@@

\subsection{Further aborted delivery attempts}

@@

\subsection{Timeouts during DATA phase}

@@

The DATA phase of the \gls{SMTP} conversation is where the headers and body
of the mail are transferred.  Sometimes there is a timeout or the
connection is lost\footnote{For brevity's sake \textit{timeout\/} will be
used throughout this section, but everything applies equally to lost
connections.} during the DATA phase; when this occurs Postfix will discard
the mail and the parser needs to discard the data associated with that
mail.  It seems more intuitive to save the mail's data to the database, but
if a timeout occurs there is no data available to save; the timeout is
recorded with the connection data instead, which is saved.

@@

\subsection{Discarding cleanup log lines}

@@

\subsection{Pickup logging after cleanup}

@@

\subsection{Smtpd stops logging}

@@

\subsection{Out of order log lines}

@@

\subsection{Yet more aborted delivery attempts}

@@

\subsection{Mail deleted before delivery is attempted}

@@

\subsection{Bounce notification mails delivered before their creation is
logged}

@@

\subsection{Mails deleted during delivery}

@@

\section{Limitations and possible improvements}

@@

    \item Each new Postfix release requires new rules to be written or
        existing rules modified to cope with the new or changed log lines.
        Similarly using a new \gls{DNSBL}, a new policy server, or new
        administrator-defined rejection messages require new rules.

@@

\subsection{Possible improvements}

>>>>

\subsection{Using a Database as an Application Programming Interface}

\label{database as API}

XXX LAST CHECKED DATES FOR ALL URLS\@.

The database populated by this program provides a simple interface to
Postfix log files.  Although the interface is a database schema rather than
a set of functions provided by a library, it performs the same function as
any other \acronym{API}: it provides a stable interface, allowing code on
either side of the interface to be changed without adverse effects, as long
as the interface is maintained.  The parser can be improved to handle
additional complications, support earlier or later releases of Postfix,
bugs can be fixed or limitations removed.  Programs that use the database
can range from the simple examples in \sectionref{motivation} to far more
complex data mining tools.

Using a database simplifies writing programs that need to interact with the
data in several ways:

    \item Facilities are provided to access databases from most programming
        languages, allowing a developer to access the data gathered using
        their preferred programming language, rather than being restricted
        to the language the parser is written in.  It is often possible to
        write an interface layer allowing code written in one language to
        be used in another language, but this greatly increases the effort
        required to use the parser.

    \item Databases provide complex querying and sorting functionality to
        the user without requiring large amounts of programming.  All
        databases provide a program, of varying complexity and
        sophistication, that can be used for ad hoc queries with minimal
        investment of time.

    \item Databases are easily extensible, e.g.:

            \item New columns can be added to the tables used by the
                program, using DEFAULT clauses or TRIGGERS to populate
                them.

            \item A VIEW gives a custom arrangement of data with minimal
                effort.

            \item If the database supports it, access can be granted on a
                fine-grained basis, e.g.\ allowing the finance department
                to produce invoices, the helpdesk to run limited queries as
                part of dealing with support calls, and the administrators
                to have full access to the data.

            \item Triggers can be written to perform actions when certain
                events occur.  In pseudo-\acronym{SQL}\@:

\begin{verbatim}
CREATE TRIGGER ON INSERT INTO results
    WHERE sender = 'boss@example.com'
        AND postfix_action = 'REJECTED'
    SEND PANIC EMAIL TO 'postmaster@example.com';
\end{verbatim}

            \item Other tables can be added to the database, e.g.\ to cache
                historical, summary, or computed data.

    \item \acronym{SQL} is reasonably standard and many people will already
        be familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn (a good
        introduction to \acronym{SQL} can be found at
        \url{http://philip.greenspun.com/sql/}, others are
        \url{http://www.w3schools.com/sql/default.asp},
        \url{http://sqlcourse.com/}).  Although every vendor implements a
        different dialect of \acronym{SQL}, the basics are the same
        everywhere (analogous to the overall similarities and minor
        differences amongst Irish English, British English, American
        English, and Australian English).  Depending on the database in use
        there may be tools available that reduce or remove the requirement
        to know \acronym{SQL}; several are available for \gls{SQLite}, the
        default database used by the parser implementation.

Storing the results in a database will also increase the efficiency of
using those results, because the log files need only be parsed once rather
than each time the data is used; indeed the results may be used by someone
with no access to the original log files.

\subsection{Rules Table}

@@

\subsection{Connections Table}

@@

    \item [server\_ip] The \acronym{IP} address (IPv4 or IPv6) of the
        server: the local address when receiving mail, the remote address
        when sending mail.

    \item [server\_hostname] The hostname of the server, it will be
        \texttt{unknown} if the \acronym{IP} address could not be resolved
        to a hostname via DNS\@.

    \item [client\_ip] The client \acronym{IP} address (IPv4 or IPv6): the
        remote address when receiving mail, the local address when sending
        mail.

    \item [client\_hostname] The hostname of the client, it will be
        \texttt{unknown} if the \acronym{IP} address could not be resolved
        to a hostname via DNS\@.

    \item [helo] The hostname used in the HELO command.  The HELO hostname
        occasionally changes during a connection, presumably because spam
        or virus senders think it is a good idea.  By default Postfix only
        logs the HELO hostname when it rejects an \acronym{SMTP} command,
        but it is quite simple to rectify this, as described in
        \sectionref{logging helo}.

@@

\subsection{Results Table}

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \acronym{SMTP} command
        --- a mechanism that is quite useful for testing new restrictions.
        This field will be 1 if the log line parsed was a warning rather
        than a real rejection, or 0 for a real rejection.  This field
        should be ignored if the result is not a rejection, i.e.\ the
        action field of the associated rule is not \texttt{REJECTION}.

    \item [smtp\_code] The \acronym{SMTP} code associated with the log
        line.  In general an \acronym{SMTP} code is only present in a log
        line for a rejection or final delivery; results initially missing
        an \acronym{SMTP} code will duplicate the \acronym{SMTP} code of
        other results in that connection.  Some final delivery log lines do
        not contain an \acronym{SMTP} code: in those cases the code is
        specified by the rule, based on the success or failure represented
        by the log line.

@@

    \item [data] A field available for anything not covered by other
        fields, e.g.\ the rejection message from a \acronym{DNSBL}\@.

@@

\subsection{Example Rule}

@@

\subsection{Creating New Rules}

@@

The log files produced by Postfix differ from installation to installation,
because administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different
\acronym{DNSBL} services, policy servers, or custom rejection messages.  To
facilitate parsing new log lines, the parser's design separates parsing
rules from parsing actions: adding new actions can be difficult, but adding
new rules to parse new rejection messages is trivial, and occurs much more
frequently.  A program is provided to ease the process of creating new
rules from unparsed log lines, based on the algorithm developed by Risto
Vaarandi for his \acronym{SLCT}~\cite{slct-paper}.  The differences between
the two algorithms will be outlined as part of the general explanation
below.

The core of the \acronym{SLCT} algorithm is quite simple: log lines are
generally created by substituting variable words into a fixed pattern, and
analysis of the frequency with which each word occurs can be used to
determine whether the word is variable or part of the fixed pattern.  This
classification can be used to group similar log lines and generate a regex
to match each group of log lines.

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained when writing rules and performs a large number of
        substitutions on the input log lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \acronym{IP} addresses, the
        standard start of rejection messages, etc.) with regex keywords
        that the parser will expand when it loads the rule (see the
        regex entry in \sectionref{rule attributes}).  The purpose of
        this step is to utilise existing knowledge to create more accurate
        regexes; it replaces a large number of variable words with fixed
        words, improving the subsequent classification of words as fixed or
        variable.  The new log lines are written to a temporary file, which
        all subsequent stages use instead of the original input file.

@@

\subsection{Actions in Detail}

@@

            \item [pid] The \acronym{pid} of the program that logged the
                line.

@@

    \item [CONNECT] Handle a remote client connecting: create a new state
        table entry (indexed by \daemon{smtpd} \acronym{pid}) and save both
        the client hostname and \acronym{IP} address.

@@

    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an
        \acronym{SMTP} command from the remote client: log the rejection
        with a mail if there is a queueid in the log line, or with the
        connection if not.

@@

\subsection{Postfix Actions}

@@

\subsection{Adding New Actions}

@@

    \item Mail enters the system via \acronym{SMTP} or local submission.

@@

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop}, by \acronym{SMTP}, by re-injection due to
forwarding, or internally generated by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately and the sender address is
logged (action: PICKUP\@; flowchart:~2).  Re-injection due to forwarding
sadly lacks explicit log lines of its own; it is explained in
\sectionref{Re-injected mails}.

@@

\acronym{SMTP} is more complicated than local submission:

@@

    \item This is followed by rejection of sender address, recipient
        addresses, client \acronym{IP} address or hostname, HELO hostname,
        etc.\ (action: REJECTION\@; flowchart:~4); acceptance of one or
        more mails (action: CLONE\@; flowchart:~5); or some interleaving of
        both.

    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
        Postfix has rejected any \acronym{SMTP} commands the data will be
        saved to the database; if not there will not be any data to save
        (any mails accepted will already have been cloned so their data is
        in another data structure).

@@

\section{Complications Encountered}

@@

\subsection{Queueid Vs Pid}

The mail lacks a queueid until it has been accepted, so log lines must
first be correlated by the \daemon{smtpd} \acronym{pid}, then transition to
being correlated by the queueid.  This is relatively minor, but does
require:

@@

    \item Most importantly: every section of code must know whether it
        needs to lookup the data structures by \acronym{pid} or queueid.

@@

\subsection{Connection Reuse}

@@

Multiple independent mails may be delivered across one connection: this
requires the algorithm to clone the current data as soon as a mail is
accepted, so that subsequent mails will not trample over each other's data.
This must be done every time a mail is accepted, as it is impossible to
tell in advance which connections will accept multiple mails.  Once the
mail has been accepted its log lines will not be correlated by
\acronym{pid} any more, its queueid will be used instead (except when
timeouts occur during the data phase
\sectionref{timeouts-during-data-phase}).  If the original connection has
any useful data (e.g.\ rejections) it will be saved to the database when
the client disconnects.  One unsolved difficulty is distinguishing between
different groups of rejections, e.g.\ when dealing with the following
sequence:

@@

    \item The client issues the RSET command to reset the \acronym{SMTP}
        session.

@@

\subsection{Re-injected Mails}

@@

    \item Local delivery re-injects the mail and logs a relayed delivery
        rather than delivering directly to a mailbox or program as it
        usually would.\footnote{Relayed delivery is performed by the
        \acronym{SMTP} client; local delivery means local to the server,
        i.e.\ an address the server is final destination for.}  In this
        case the mail may already have been created (described above) and
        the unknown origin flag will be cleared; if not a new data
        structure will be created.  In both cases the re-injected mail is
        marked as a child of the original mail.  The log line in question
        is:

@@

\subsection{Identifying Bounce Notifications}

@@

    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
        messages associated with the mail, indicating it was generated
        internally by Postfix, not accepted via \acronym{SMTP} or submitted
        locally by \daemon{postdrop}.

@@

\subsection{Aborted Delivery Attempts}

@@

Some mail clients behave unexpectedly during the \acronym{SMTP} dialogue:
the client aborts the first delivery attempt after the first recipient is
accepted, then makes a second delivery attempt for the same recipient which
it continues with until delivery is complete.  Microsoft Outlook is one
client that behaves in this fashion; other clients may act in a similar
way.  An example dialogue exhibiting this behaviour is presented below
(lines starting with a three digit number are sent by the server, the other
lines are sent by the client):

@@

If all the checks above are successful then the mail is assumed to be an
aborted delivery attempt and is discarded.  There will be no further
entries logged for such mails, so without identifying and discarding them
they accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and \acronym{IP} address, but the database
schema requires many more fields be populated (see \sectionref{connections
table} and \sectionref{results table}).  This heuristic is quite
restrictive, and it appears there is little scope for false positives; if
there are any false positives there will be warnings when the next log line
for that mail is parsed.  False negatives are less likely to be detected:
there may be queueid clashes (and warnings) if mails remain in the state
tables after they should have been removed, otherwise the only way to
detect false negatives is to examine the state tables after each parsing
run.

@@

\newpage{} % XXX

\subsection{Further Aborted Delivery Attempts}

@@

\subsection{Timeouts During DATA Phase}

@@

The DATA phase of the \acronym{SMTP} conversation is where the headers and
body of the mail are transferred.  Sometimes there is a timeout or the
connection is lost\footnote{For brevity's sake \textit{timeout\/} will be
used throughout this section, but everything applies equally to lost
connections.} during the DATA phase; when this occurs Postfix will discard
the mail and the parser needs to discard the data associated with that
mail.  It seems more intuitive to save the mail's data to the database, but
if a timeout occurs there is no data available to save; the timeout is
recorded with the connection data instead, which is saved.

@@

\subsection{Discarding Cleanup Log Lines}

@@

\subsection{Pickup Logging After Cleanup}

@@

\subsection{Smtpd Stops Logging}

@@

\subsection{Out of Order Log Lines}

@@

\subsection{Yet More Aborted Delivery Attempts}

@@

\subsection{Mail Deleted Before Delivery is Attempted}

@@

\subsection{Bounce Notification Mails Delivered Before Their Creation Is
Logged}

@@

\subsection{Mails Deleted During Delivery}

@@

\section{Limitations and Possible Improvements}

@@

    \item Each new Postfix release requires new rules to be written or
        existing rules modified to cope with the new or changed log lines.
        Similarly using a new \acronym{DNSBL}, a new policy server, or new
        administrator-defined rejection messages require new rules.

@@

\subsection{Possible Improvements}

### 2009-02-12 doc/logparser-introduction.tex
Fixed acronyms so they work properly, get expanded when appropriate, and add the     correct entries to the glossary and acronym appendices. Correct capitalisation of most section and subsection headings. Corrections made to background chapter; more remain to be made.

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check if new anti-spam measures are working.  The more
adventurous will have generated statistics to show how successful each of
their anti-spam measures has been in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the author's first real foray into processing
Postfix log files.}  Very few will have performed in-depth parsing and
analysis of their log files, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing log lines
independently.  One of the barriers to this kind of processing is the
unstructured nature of Postfix log files, where each log line was added on
an ad hoc basis as a requirement was discovered or new functionality was
added.\footnote{A history of all changes made to Postfix is distributed
with the source code, available from \url{http://www.postfix.org/}} Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every
\gls{DNSBL}\footnote{This document is supplied with a glossary, see
\textsection\ref{Glossary}.} returns a different explanatory message;
policy servers may log different messages depending on the characteristics
of the connection.  There are many ways in which the log lines may differ
between servers, even within the same organisation: servers may be
configured differently, or running different versions of Postfix.  This
paper documents the difficult process of parsing Postfix log files,
presenting \gls{PLP}, a program that parses Postfix log files and places
the resulting data into a database for later use.  The gathered data can
then be used to optimise current anti-spam measures, provide a baseline to
test new anti-spam measures against, or to produce statistics showing how
effective those measures are.  Numerous other uses are possible for such
data: improving server performance by identifying troublesome destinations
and reconfiguring appropriately; identifying regular high volume uses
(e.g.\ customer newsletters) and restricting those uses to off-peak times;
detecting virus outbreaks that propagate via email; as a base for billing
customers on a shared server.  Preserving the raw data enables users to
develop a multitude of uses far beyond those conceived of by the author.

@@

Section~\ref{background} provides background information useful in
understanding the thesis, parser, and architecture.  It introduces the idea
of using a database schema as an \gls{API}, to provide a language-neutral
interface to the data gathered.  The novel separation of rules, actions,
and framework is discussed, giving the reasons that approach was taken when
designing the parser.

@@

The bibliography contains references to the resources used in developing
the algorithm, writing the program, and preparing this thesis.  Also
listed are some additional resources expected to be helpful in
understanding \gls{SMTP}, Postfix, anti-spam techniques, or the thesis.

>>>>

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check if new anti-spam measures are working.  The more
adventurous will have generated statistics to show how successful each of
their anti-spam measures has been in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the author's first real foray into processing
Postfix log files.}  Very few will have performed in-depth parsing and
analysis of their log files, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing log lines
independently.  One of the barriers to this kind of processing is the
unstructured nature of Postfix log files, where each log line was added on
an ad hoc basis as a requirement was discovered or new functionality was
added.\footnote{A history of all changes made to Postfix is distributed
with the source code, available from \url{http://www.postfix.org/}} Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every
\acronym{DNSBL}\footnote{This document is supplied with a glossary, see
\textsection\ref{Glossary}.} returns a different explanatory message;
policy servers may log different messages depending on the characteristics
of the connection.  There are many ways in which the log lines may differ
between servers, even within the same organisation: servers may be
configured differently, or running different versions of Postfix.  This
paper documents the difficult process of parsing Postfix log files,
presenting \parsername{}, a program that parses Postfix log files and
places the resulting data into a database for later use.  The gathered data
can then be used to optimise current anti-spam measures, provide a baseline
to test new anti-spam measures against, or to produce statistics showing
how effective those measures are.  Numerous other uses are possible for
such data: improving server performance by identifying troublesome
destinations and reconfiguring appropriately; identifying regular high
volume uses (e.g.\ customer newsletters) and restricting those uses to
off-peak times; detecting virus outbreaks that propagate via email; as a
base for billing customers on a shared server.  Preserving the raw data
enables users to develop a multitude of uses far beyond those conceived of
by the author.

@@

Section~\ref{background} provides background information useful in
understanding the thesis, parser, and architecture.  It introduces the idea
of using a database schema as an \acronym{API}, to provide a
language-neutral interface to the data gathered.  The novel separation of
rules, actions, and framework is discussed, giving the reasons that
approach was taken when designing the parser.

@@

The bibliography contains references to the resources used in developing
the algorithm, writing the program, and preparing this thesis.  Also
listed are some additional resources expected to be helpful in
understanding \acronym{SMTP}, Postfix, anti-spam techniques, or the thesis.
